Pre-processing high frequency animal tracking data

### 2020-08-25 02_filtering_data.Rmd
rename and add work on section to compare

```{r prep_libs_02_01}
library(data.table)
library(atlastools)
```

@@

```{r read_sim_data}
# read in the data
data <- fread("data/data_sim.csv")

# add some gross positioning errors
data[sample(nrow(data), 100), c("x", "y") := lapply(.SD,
                                                   function(z) {
                                                     z + rnorm(100, sd = 0.1)
                                                   }),
     .SDcols = c("x", "y")]

# add a fictional NBS covariate
data[, NBS := floor(rnorm(n = nrow(data), mean = 4, sd = 1))]
```

## Excluding gross positioning errors

@@

```{r}
# remove positions outside a bounding box
data_inside_bbox <- data.table::copy(data)

@@

## Excluding data on covariate values

Some position covariates may be indicators of unreliable data, due to being linked with increased spatial uncertainty. The number of base-stations (NBS) used to localise a transmitting tag in ATLAS systems is one such covariate, with a minimum of three base-stations required for a two-dimensional localisation. Here, we remove all positions with < 3 base-stations.

```{r}
# remove positions with few base stations
data_many_nbs <- copy(data_inside_bbox)
data_many_nbs <- atl_filter_covariates(data_many_nbs, 
                                       filters = c("NBS >= 3"))
```

## The non-movement approach

Bjorneraas et al. (2010) describe intuitive filtering steps for trajectories that are based in the biology of the study species (moose _Alces alces_ in their case). Briefly, this 'non-movement method' identifies implausible segments of a trajectory, and removes them. Here, we adopt elements of their method to identify and remove 'spikes' in the movement track, i.e., positions which indicate an improbably fast and short excursion away from the real trajectory.

@@

```{r}
# get speed and turning angle
data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs),
                     out_speed = c(atl_get_speed(data_many_nbs)[-1], NA),
                     angle = atl_turning_angle(data_many_nbs))]

@@

# filter the copy
data_non_move <- atl_filter_covariates(data_non_move,
                                       filters = c("in_speed < 0.0012",
                                                   "out_speed < 0.0012",
                                                   "angle < 40"))
```

@@

A curious issue affecting TOA tracking systems is the phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.

@@

A reflection appears thus:

@@

# add a reflection
data_with_reflection[500:600, c("x", "y") := runif(101, min = 1, max = 1.1)]
```

@@

```{r}
# attempt to remove reflections
data_no_reflections <- copy(data_with_reflection)

data_no_reflections <- atl_remove_reflections(data_with_reflection,
                                              point_angle_cutoff = 40,
                                              reflection_speed_cutoff = 0.0012)
```

>>>>

```{r prep_libs_02_01}
# to handle movement data
library(data.table)
library(atlastools)

# to simulate movement
library(smoove)

# to generate a landscape
library(raster)
library(NLMR)

# to plot
library(ggplot2)
```

@@

```{r read_sim_data}
# read in the data
data <- fread("data/data_sim.csv")
```

```{r function_add_outliers}
# function for point outliers
do_add_outliers <- function(data, p_data = 0.1, std_dev = 0.05) {
  n_rows <- floor(p_data * nrow(data))
  tmp_data <- copy(data)

  tmp_data[sample(nrow(data), n_rows),
            c("x", "y") := lapply(.SD,
                                  function(z) {
                                    z + stats::rnorm(n_rows, sd = std_dev)
                                  }),
            .SDcols = c("x", "y")]
  return(tmp_data)
}
```

We add outliers at random to the data to demonstrate their removal.

```{r add_outlier}
# add 100 outliers
data_copy <- do_add_outliers(data, p_data = 0.01, std_dev = 0.05)
```

## Removing gross positioning errors

@@

```{r remove_outside_bbox, eval=FALSE}
# remove positions outside a bounding box
data_inside_bbox <- data.table::copy(data_copy)

@@

```{r}
ggplot()+
  geom_path(data = data_copy, aes(x, y), col = "red")+
  geom_path(data = data_inside_bbox, aes(x, y))
```

## Removing point outliers: The non-movement approach

Bjorneraas et al. (2010) describe intuitive filtering steps for trajectories that are based in the biology of the study species (moose _Alces alces_ in their case). Briefly, this 'non-movement method' identifies implausible segments of a trajectory, and removes them. Here, we adopt elements of their method to identify and remove 'spikes' or point outliers in the movement track, i.e., positions which indicate an improbably fast and short excursion away from the real trajectory.

@@

```{r example_remove_outliers, eval=FALSE}
# get speed and turning angle
data_many_nbs[, `:=`(in_speed = atl_get_speed(data_many_nbs,
                                              type = "in"),
                     out_speed = atl_get_speed(data_many_nbs,
                                               type = "out"),
                     angle = atl_turning_angle(data_many_nbs))]

@@

# filter the copy
data_non_move <- atl_filter_covariates(data_non_move,
                                       filters = c("in_speed < 0.001",
                                                   "out_speed < 0.001",
                                                   "angle < 40"))
```

```{r}
ggplot()+
  geom_path(data = data_inside_bbox,
            aes(x, y),
            colour = "black")+
  geom_path(data = data_non_move,
            aes(x, y),
            colour = "red")
```

## Does removing point outliers help?

### Making error combinations

Point outliers in a trajectory can be boiled down to two parameters: (1) how many positions have been converted into point outliers or spikes, and (2) how large the spikes are relative to the scale of the movement track.

Here we create combinations of these parameters and create 25 replicates of each combination using the error generator on the imported simulated data.

```{r outlier_error_combinations}
# define unique values
p_outlier <- c(0.01, 0.05, 0.1, 0.25)
ext_outlier <- c(0.01, 0.05, 0.1)
replicate <- seq_len(1)

# make combinations
param_c <- CJ(p_outlier, ext_outlier, replicate)
```

### Procedural landscape generation

Next we procedurally generate (simulate) a landscape from which to extract habitat choice.

```{r make_landscape}
# use NLMR
landscape <- NLMR::nlm_randomcluster(100, 100, 
                                     p = 0.2, 
                                     ai = c(0.2, 0.3, 0.5),
                                     rescale = TRUE)

# save landscape as raster
raster::writeRaster(landscape,
                    filename = "data/data_landscape.tif",
                    format = "GTiff", 
                    overwrite = TRUE)
```

```{r read_landscape}
# read landscape
landscape <- raster("data/data_landscape.tif")

# scale landscape?
bounds <- c(0, 1, 0, 1)
bound_scale <- c(1, 2, 5, 10)
extent(landscape) <- bounds
```

### Make movement track replicates

```{r}
sim_data <- mapply(function(p_outlier, ext_outlier) {
  # make simulated data
  this_data <- do_smoove_data()

  # extract 'true' movement metrics
  speed <- atl_get_speed(this_data, type = "in")

  # extract proportion of landscape used
  land_vals <- extract(landscape,
                       this_data[, c("x", "y")])

  land_vals <- as.data.table(table(landscape))

  # what to return

  # WORK IN PROGRESS HERE

  error_data <- do_add_outliers(this_data, 
                                p_data = p_outlier,
                                std_dev = ext_outlier)
  return(error_data)
}, 
param_c$p_outlier, 
param_c$ext_outlier, 
SIMPLIFY = FALSE)
```

@@

A peculiar issue affecting TOA tracking systems is the phenomenon of ‘reflected’ positions, in which the estimated location is instantaneously displaced many hundreds of metres from the individual’s real position for some time, followed by a return to the real position.

@@

A reflection looks like this:

@@

# add a reflection
data_with_reflection[500:600, `:=`(x = runif(101, 1, 1.01),
                                   y = runif(101, 1, 1.01))]
```

@@

```{r}
# attempt to remove reflections
data_no_reflection <- copy(data_with_reflection)

data_no_reflection <- atl_remove_reflections(data_with_reflection,
                                              point_angle_cutoff = 10,
                                              reflection_speed_cutoff = 0.0015)
```

```{r}
ggplot()+
  geom_path(data = data_with_reflection,
            aes(x, y))+
  geom_path(data = data_no_reflection,
            aes(x, y),
            colour = "red")
```

## Assessing whether filtering helps
